Trauma treatment delay

### 2024-01-09 manuscriptlongversion.Rmd
possible final draft

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less conscious patient is more
prone to receiving delayed treatment. Compared to a previous study
analysing the same data but with OFI as the outcome, this result is
similar to what they found using the same database (27). As the
mentioned study suggests, this could be explained by that lower GCS as
well as higher ISS indicates severe trauma. More injured patients often
require more complex management and more interventions. More
interventions leave more room for delay to these interventions, why they
are more prone to delay in treatment. Respiratory rate also had the
correlation that lower respiratory rate correlated with lower rates of
delayed treatment among the patients. As the interval is \>10 to \<29 in
the triage guideline, both too low and too high respiratory rate is not
desirable(28). This might have a similar explanation to ISS and GCS,
that patients with tachypnea might receive higher rates of delayed
treatment because they are more severely injured, therefore a higher
risk of missing a possible injury. The previous study also found this,
but the correlation disappeared when adjusted for, which was
hypothesised being due to its shared variability with the ISS
variable(27). In this study, this was the case for number of injuries,
with it and ISS often being linked, as a more severely injured patient
will both score higher on ISS and number of injuries. Therefore, we
hypothesise that the effect of number of injuries becomes insignificant
when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, as many are analysing mortality.
Some similarities exist with which patient level factors predict the
outcome, with higher mortality being seen in patients with lower GCS and
higher ISS [(29)](30). Increased age was also a factor linked to
mortality, which was not seen to be linked to delayed treatment in this
study. An older patient is more likely to have comorbidities that might
affect their mortality but not increase their likelihood of receiving
delayed treatment the same way ISS and GCS does(31).

It is important to keep in mind that delay in treatment does not have a
set limit of time as a where under it is within the accepted timeframe
and if you go over it, it is considered delayed treatment. The audit
filters flag for a wait for CT over 30 minutes and a wait for operation
over 60 minutes, which then gets reviewed by previously mentioned two
specialised nurses. However, if a patient has a longer wait, it does not
necessarily mean that there was a delay. If it takes a longer time for a
trauma patient to receive a CT or operation, it might be because there
were more urgent matters who had to be taken care of, alternatively, the
MoM conference might deem it as the delay was justified for other
reasons. The point being, it is clear that the mean waiting time for CT
and operation is higher than 30 and 60 minutes respectively when looking
at the database. Using this audit filter as the definition for delay
would not mean anything, as the majority of patients would be classified
as having waited for too long. Instead, it is the MoM conferences that
decide if the delay had any adverse effects for the patient, such as an
injury worsening during the time spent waiting without a clear
motivation to why it had to wait. It is these cases that they label it
as the "too long until CT '' or"Too long until operation". Meaning, all
delays in the database are actual assessed delays, not just a time limit
that has been surpassed.

The results of the study are expected as previous similar studies have
come to similar conclusions. A goal to strive towards in trauma care
through research and improvement programs is that no patient level
factors should be linked with a higher risk of delayed treatment. The
variables found in this study to be linked with delayed treatment are
however factors that originate from the trauma, not previously existing
in the patient such as age or gender. This implies that patients are not
systematically discriminated against based on their gender or age, which
is a positive sign seen from a health equity point of view.

@@

In this study, one of the limitations was the number of people having
the recorded outcome being quite small, leading us into having to limit
the number of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories, from 1-4. However, many of the RTS variables had few to no
patients with the observed outcome. This led to separation in the
logistic regression in these variables, referring to when an outcome
variable separates a predictor variable completely, which occurs when
the predictor is only associated with one outcome value (32). In the
case of these variables, it was only RTS 3 and 4 who had patients with
the observed outcome, which left RTS 1 and 2 with confidence intervals
of 0.00 to infinity. This way would have been the best way to examine
these variables, but as that was not possible with this group of
patients, we resorted to making these variables continuous. We
considered expanding the outcome group to include more OFI, but opted
for the solution used in the study instead as we felt that including
more OFI would lead to less clinically applicable results.

The problem mentioned above with too few patients in the outcome group
made this into a post hoc analysis. As we did not know the number of
people who received delayed treatment when we handled the scrambled
data, we ran the code on the real data when we had too many variables
for the dataset. Therefore, we had to change the variables after having
seen the real data, making it a post hoc analysis. This could in theory
open up for possible selection bias, as we might choose variables which
give us results we are looking for. However, all the variables included
in the final analysis were already included in the first one. Variables
were only removed, not added after having seen the real data. The best
way to have gone about this would have been to just look at the amount
of people who received delayed treatment before choosing the variables.
Then, choosing an amount of variables that would work with the amount of
patients. This would have avoided making it post hoc and changing the
variables after seeing the data.

On discussing possible biases, misclassification bias might also be
present in this study, though it is something that we cannot control in
this study. As previously mentioned, the possible existence of an OFI is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialised
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

@@

Further analysing the results and its clinical applications, the results
can be interpreted as that it is mostly injured patients with affected
consciousness that receive delayed treatment. The clinical applications
of this can be seen as we need to do more to avoid this delay for this
patient group. It is important to remember that the delays are all
patients that have received adverse outcomes because of their delay.
Some might include injuries being missed or not being treated. The study
highlights the urgency in these patients and the heightened risk for
delay they face due to their situation. If clinicians are aware that
these patient factors receive delayed treatment more frequently, active
steps can be taken towards reducing this risk and bettering the trauma
care.

This also highlights the importance of adequate diagnostic imaging when
the patient is stabilised. It occurs that patients are stabilised in the
operating room (OR), gets taken to the CT then back to the OR. This is
important as to not miss any injury, especially in severely injured
patients where it can be difficult to identify all injuries without
diagnostic imaging. When clinicians have access to these instruments
they should be used whenever deemed necessary, as studies suggest it has
a positive effect on patient outcome [@Hajibandeh_2014].

Further applying the study clinically, it is also important to educate
clinicians on what steps went wrong in these patients and what could
have been done to prevent it. If clinicians know the most common
mistakes made in this patient group, they might be less likely to make
these mistakes themselves, possibly lowering the risk of receiving
delayed treatment for these patients with these factors. The results of
this study and similar ones should be communicated to trauma centers, as
the goal is to always better the trauma care for patients.

@@

Discussing the generalisability of this study, we included all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
ofi. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. While the population only represents Karolinska
university hospital, the conclusion should be applicable to the whole
trauma population.

@@

In future studies, various methods to further examine the association
between delayed treatment and various variables, alternatively using
another OFI as the outcome variable. Examining different variables such
as highest hospital care level, intubation status or first emergency
procedure performed might be interesting variables, among others. To be
able to do this, more patients in the outcome group would be needed,
either by expanding the register to several hospitals, or expanding the
outcome group by adding different ofi. This however would make the study
less specific for delayed treatment. Further studies might also look
into different ofi other than delayed treatment, such as missed injury,
faulty communication or triage.

@@

In conclusion, based on this single trauma center, the results suggest
that it is mostly more injured and patients with affected consciousness
that receive delayed treatment. Clinicians should be aware of this as
they can take steps to lower that risk for this patient group. Further
studies need to be conducted on this subject to further examine the link
between delayed treatment and other patient level factors.

>>>>

The aim was to determine the associations between patient level factors
and receiving delayed treatment in adult trauma patients. A logistic
regression on seven variables showed that higher ISS and RR and lower
GCS was significantly associated with patients receiving delayed
treatment.

This suggests that a heavily injured and less conscious patient is more
prone to receiving delayed treatment. Compared to a previous study
analysing the same data but with OFI as the observed outcome, this
result is similar to what they found[@Pat_factors]. As the mentioned
study suggests, this could be explained by that lower GCS as well as a
higher ISS indicates severe trauma. More injured patients often require
more complex management and a higher number of interventions, increasing
the risk of delayed treatment.

Respiratory rate had the correlation that a lower respiratory rate was
associated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This could have a
similar explanation to ISS and GCS, that patients with tachypnea might
receive higher rates of delayed treatment because they are more severely
injured, therefore a higher risk of missing a possible injury. The
previous study also found this, but the correlation disappeared when
adjusted for, which said study hypothesised being due to its shared
variability with the ISS variable[@Pat_factors]. In this study, the same
correlation could be seen with NOI. A more severely injured patient will
score higher on both ISS and NOI. Therefore, our thought is that the
effect of NOI becomes insignificant when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, as many are analysing mortality.
Some similarities exist between which patient level factors predict the
outcome, with higher mortality being seen in patients with lower GCS and
higher ISS [@Mort_sammy][@Mort_hefny]. Advanced age has also been
identified as a contributory factor to increased mortality. However, it
is noteworthy that this study did not establish a significant
association between age and delayed treatment. Unlike ISS and GCS, which
are indicative of the severity of injuries and the immediate treatment
needs, advanced age is more likely to introduce comorbidities that
affect overall mortality but may not necessarily heighten the
probability of receiving delayed treatment[@Gioffre2018].

When discussing the outcome delay in treatment used in this study, it is
not a set time limit where if the wait exceeds it, it is considered
delayed treatment. The audit filters used at Karolinska University
Hospital flag for a wait for CT over 30 minutes and a wait for operation
over 60 minutes, which then gets reviewed by previously mentioned two
specialised nurses. However, if a patient has a longer wait, it does not
necessarily mean that there was a delay. 30 and 60 minutes are time
limits set to have a high sensitivity and low specificity in the audit
filters. Whether it is a delay is decided by the MoM conferences. It is
clear that the mean waiting time for CT and operation is higher than 30
and 60 minutes respectively when looking at the database. Using this
audit filter as the definition for delay would not mean anything, as the
majority of patients would be classified as having waited for too long.
Instead, the MoM conferences decide if the wait had any adverse effects
on the patients, such as an injury worsening during this time. It is
these cases that they label it as the \"too long until CT\" or \"Too
long until operation\". All delays in the database are delays that
adversely affected the patients.

@@

In this study, one of the limitations was the number of patients having
the recorded outcome being quite small, leading us into having to limit
the number of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories from 1-4, since this would have allowed for a more detailed
analysis. However, the sample size was too small in the outcome group
which led us to resort to making these variables continuous. We
considered expanding the outcome group to include more OFI, but opted
for the solution used in the study instead as we felt that including
more OFI would lead to less clinically applicable results.

The problem mentioned above with too few patients in the outcome group
made the study into a post hoc analysis. We were not aware of the number
of patients who received delayed treatment when we handled the scrambled
data as we wanted to minimise selection bias. When we ran the code on
the real data it became clear that the sample size was too small for our
approach. Therefore, we had to change the variables after having seen
the real data, making it a post hoc analysis. This could in theory open
up for selection bias, as we might have chosen variables which give us
results we are looking for. We did our best to remove variables at
random, but it is possible that we subconsciously removed variables that
would change the result in a way benefiting us. The best way to have
gone about this would have been to calculate the amount of people who
received delayed treatment before choosing the variables. Then, choosing
an amount of variables that would work with the amount of patients. This
would have avoided making it post hoc and minimised selection bias.

On discussing possible biases, misclassification bias might also be
present in this study, though it is something that we cannot control in
this study. As previously mentioned, the possible existence of an OFI is
decided during the MoM conferences. However, they might not always be
correct in their judgement and occasionally misclassify, leading to
patients being in the wrong group in the analysis. Also, some patients
might have been missed even before reaching the MoM conference. The
patients included in the conference are selected through audit filters
and two specialised nurses, and patients might be missed in this stage.
This bias should not be big enough to have a significant impact on the
results, but can be kept in mind.

@@

Analysing the results and its clinical applications, the results can be
interpreted as that it is mostly injured patients with affected
consciousness that receive delayed treatment. The clinical applications
of this are that clinicians need to do more to avoid this delay for this
patient group. All patients in the outcome group are patients that
experienced adverse effects because of their delayed treatment. Some
might include injuries being missed or not being treated, leading to
unnecessary suffering for the patient. Clinicians being aware of the
patient level factors at risk for delayed treatment might help with
their decision making when caring for these patients.

It is important to highlight the importance of adequate diagnostic
imaging when the patient is stabilised, as a substantial part of the
patients in the outcome group were patients who waited too long for
diagnostic imaging. To lower this risk, it occurs that patients are
stabilised in the operating room (OR), gets taken to the CT then back to
the OR. This is done to minimise the risk of missing an injury,
especially in severely injured patients where it can be difficult to
identify all injuries without diagnostic imaging. When clinicians have
access to these instruments they should be used whenever deemed
necessary, as studies suggest it has a positive effect on patient
outcome [@Hajibandeh_2014].

#### Equity

The results of the study are expected as previous similar studies have
come to similar conclusions. A goal to strive towards in trauma care
through research and improvement programs is that no patient level
factors should be linked with a higher risk of delayed treatment. The
variables found in this study to be linked with delayed treatment are
patient factors that arise from the experienced trauma, not previously
existing in the patient such as age or gender. This implies that
patients are not systematically discriminated against based on their
gender or age, which is a positive sign seen from a health equity point
of view.

@@

Discussing the generalisability of this study, we included all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
OFI. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. While the population only represents Karolinska
university hospital, the conclusion should be applicable to the whole
trauma population.

@@

To further expand the knowledge on what patient level factors are
associated with delayed treatment, future studies should aim for a
larger sample size to be able to analyse more variables, such as highest
hospital care level, intubation status or first emergency procedure
performed. To be able to do this, more patients in the outcome group
would be needed. This can be done by conducting a study which includes
the registries of several hospitals.

A question that arose during this study was what factors are associated
with other OFI, such as missed injury, communication errors or faulty
triage, and if they are similar to the ones found in this study. Future
studies could use the same data but use another OFI as the outcome to
compare how variables associated with each OFI differ.

@@

Based on this single trauma center, the results suggest that it is
mostly more injured and patients with affected consciousness that
receive delayed treatment. Clinicians should be aware of this as they
can take steps to lower that risk for this patient group. Further
studies need to be conducted on this subject to further examine the link
between delayed treatment and patient level factors.
